Determining the existence and direction of causality in the face of measurement error

### 2016-12-14 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Determining the existance and direction of causality in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{13 December 2016}

@@

Simulations were conducting by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_O\), the
outcome \(y\), the measured values of the outcome \(y_O\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_O = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
in the measurement of \(x\). Here the true value of the exposure is
partially explained by the genetic instrument, \(g\), such that

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement
error is introduced it can be shown using basic covariance properties
(Appendix 1) that

\[
\begin{aligned}
cov(g, y - \hat{y}) & = cov(g, y_O) - cov(g, \hat{y}_O)  \\
                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
\end{aligned}
\]

@@

Thus an observational study will find \(cov(g, y_O - \hat{y_O}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_O | x\), which is in violation of the the 4th
condition of the CIT. Note that linear transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions.

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the apparent evidence for
the incorrect inference.

@@

\[
\begin{aligned}
\beta_{MR} = \frac{\beta_{y_O}}{\beta_g} & = \frac{cov(y_O, g)}{cov(x_O, g)} \\
                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
\end{aligned}
\]

Thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
scaled by the measurement error in \(x\) and \(y\), but is unrelated to
measurement imprecision. Measurement error and imprecision are likely to
reduce the power of MR inferring a causal association (increased false
negative rate), but importantly the evidence for causal association,
based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
inflated when the null hypothesis is true.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 2 shows that when there is no measurement error
in \(y\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(x\). It also shows
that in most cases where \(y\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
condition of \(d > 0\) is satisfied.

@@

\begin{longtable}[]{@{}lrrrr@{}}
\toprule
Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
\(cor(\rho_{MR}, \rho_{P})\)\tabularnewline
\midrule
\endhead
Expression causes Methylation & 246 & 162 & 0.3902439 &
0.5629453\tabularnewline
Methylation causes Expression & 212 & 110 & 0.2877358 &
0.4384660\tabularnewline
\bottomrule
\end{longtable}

@@

Figure 2: Plots depicting the parameter space in which the function
\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
negative the Steiger test is liable to infer the wrong direction of
causality. \(d\) is plotted on the y axis, with measurement precision of
\(x\) (\(cor(x,x_O)\)) on the x-axis. Higher measurement precision of
\(y\) (\(cor(y,y_O)\)) is depicted by lighter coloured lines, and
columns of boxes are for different true values of \(cor(x,y)\). Shaded
regions show the parameter space where \(d\) is negative. The graph
shows that for the majority of the parameter space of the function,
\(d\) is positive, especially where causal relationships are relatively
weak.

@@

Figure 4: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations. For the CIT method, outcome 1 denoted
evidence for causality with correct model, outcomes 2 or 3 denoted
evidence for causality with incorrect model, and outcome 4 denoted no
evidence for causality.

@@

\[
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), \(x_O\) is the measured quantity of
\(x\), where measurement error is incurred from linear transformation in
\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
\(\epsilon_{mx}\), and \(y_O\) is the measured quantity of \(y\), where
measurement error is incurred from linear transformation in
\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
of association between \(g\) and \(y\) after conditioning on \(x\).
Under the CIT, this is expected to be \(cov(g, y_O - \hat{y}_O) = 0\)
when \(x\) causes \(y\), where
\(\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O\) is the predicted
value of \(y_O\) using the measured value of \(x_O\).

We can split \(cov(g, y_O - \hat{y}_O)\) into two parts, \(cov(g, y_O)\)
and \(cov(g, \hat{y}_O)\).

@@

\[
\begin{aligned}
cov(g, y_O) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_O} \beta_{mx} \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}_O) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, y_O - \hat{y}_O) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

thus \(cov(g, y_O - \hat{y}_O) = 0\) if the measurement imprecision in
\(x_O\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
then the condition \(cov(g, y_O - \hat{y}_O) = 0\) will not hold.

@@

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1). In the presence of measurement error in \(x\) and \(y\),
however, Steiger test will instead be assessing the inequality
\(\rho_{g, x_O} > \rho_{g, y_O}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
\]

@@

Sensitivity analysis for the Steiger test involves evaluating the
probability that the inferred causal direction is incorrect due to
differences in measurement error between the hypothesised exposure,
\(x\), and hypothesised outcome, \(y\), variables. The objective is to
calculate what the true \(\rho_{g, x}\) and \(\rho_{g, y}\) values would
be given empirical estimates of \(\rho_{g, x_O}\) and \(\rho_{g, y_O}\)
across all possible values of measurement error.

\[
\rho_{g,x} = \frac{\rho_{x,x_O}}{\rho_{g,x_O}}
\]

\[
\rho_{g,y} = \frac{\rho_{y,y_O}}{\rho_{g,y_O}}
\]

If \(\rho_{g,x} > \rho_{g,y}\) then we infer that the causal direction
is from exposure to outcome. Assume that all measurement error values
for \(x\) and \(y\) are equally probable, the surface of possible values
of this inequality is
\(z = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
and the objective is to calculate \(P(z > 0)\), which corresponds to the
proportion of measurement error values that would lead to the Steiger
test being incorrect.

The surface area of \(z\) across the domain
\(\rho_{g,y_O} \leq \rho_{y,y_O} \leq 1; \rho_{g,x_O} \leq \rho_{x,x_O} \leq 1\)
is obtained from

\[
\begin{aligned}
S_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{  \frac{\partial}{\partial \rho_{x,x_o}} \left (\frac{\rho_{g,x_O}}{\rho_{x,x_O}} \right )^2 +  \left (\frac{\rho_{g,y_O}}{\rho_{y,y_O}} \right )^2   +1 }\, dydx \\
& = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
\end{aligned}
\]

and the proportion of the surface for which \(z > 0\) is obtained from

\[
\begin{aligned}
S_{z>0} = \int^1_{\rho_{g,x_o}} \int^{R}_{\rho_{g,y_o}} \sqrt{ \frac{\rho_{g,x_O}^2}{\rho_{x,x_O}^4} + \frac{\rho_{g,y_O}^2}{\rho_{y,y_O}^4} + 1}\, dydx
\end{aligned}
\]

where \(R\) is the function for which
\(0 = \frac{\rho_{y,y_O}}{\rho_{g,y_O}} - \frac{\rho_{x,x_O}}{\rho_{g,x_O}}\),
i.e.

\[
R = \frac{\rho_{g,y_o}\rho_{g,x}}{\rho_{g,x_o}}
\]

Finally, the probability that the inference of \(x \rightarrow y\)
obtained from the Steiger test is incorrect due to measurement error is
obtained from

\[
P(M) = \frac{S_{z>0}}{S_z}
\]

\(P(M)\) can range from 0 to 1, where a small value indicates that the
Steiger test is relatively insensitive to measurement error. This value
is estimated through simulation.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Determining the existence and direction of causality in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existence and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{14 December 2016}

@@

Simulations were conducting by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_o\), the
outcome \(y\), the measured values of the outcome \(y_o\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_o = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
in the measurement of \(x\). Here the true value of the exposure is
partially explained by the genetic instrument, \(g\), such that

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
error is introduced it can be shown using basic covariance properties
(Appendix 1) that

\[
\begin{aligned}
cov(g, y - \hat{y}) & = cov(g, y_o) - cov(g, \hat{y}_o)  \\
                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
\end{aligned}
\]

@@

Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_o | x\), which is in violation of the the 4th
condition of the CIT. Note that linear transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions.

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the apparent evidence for
the incorrect inference.

@@

\[
\begin{aligned}
\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
\end{aligned}
\]

Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
scaled by the measurement error in \(x\) and \(y\), but is unrelated to
measurement imprecision. Measurement error and imprecision are likely to
reduce the power of MR inferring a causal association (increased false
negative rate), but importantly the evidence for causal association,
based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
inflated when the null hypothesis is true.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 2 shows that when there is no measurement error
in \(y\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(x\). It also shows
that in most cases where \(y\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
condition of \(d > 0\) is satisfied.

@@

\begin{longtable}[]{@{}lrrrrr@{}}
\toprule
Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
\(cor(\rho_{MR}, \rho_{P})\) & Mean sensitivity\tabularnewline
\midrule
\endhead
Expression causes Methylation & 246 & 162 & 0.3902439 & 0.5629453 &
0.2031556\tabularnewline
Methylation causes Expression & 212 & 110 & 0.2877358 & 0.4384660 &
0.2365816\tabularnewline
\bottomrule
\end{longtable}

@@

Figure 2: Plots depicting the parameter space in which the function
\(d = cor(x, x_o) - cor(x,y)cor(y, y_o)\) is negative. When \(d\) is
negative the Steiger test is liable to infer the wrong direction of
causality. \(d\) is plotted on the y axis, with measurement precision of
\(x\) (\(cor(x,x_o)\)) on the x-axis. Higher measurement precision of
\(y\) (\(cor(y,y_o)\)) is depicted by lighter coloured lines, and
columns of boxes are for different true values of \(cor(x,y)\). Shaded
regions show the parameter space where \(d\) is negative. The graph
shows that for the majority of the parameter space of the function,
\(d\) is positive, especially where causal relationships are relatively
weak.

@@

Figure 4: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations. For the CIT method, outcome 1 denoted
evidence for causality with correct model, outcomes 2 or 3 denoted
evidence for causality with incorrect model, and outcome 4 denoted no
evidence for causality.

@@

\[
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), \(x_o\) is the measured quantity of
\(x\), where measurement error is incurred from linear transformation in
\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
measurement error is incurred from linear transformation in
\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
of association between \(g\) and \(y\) after conditioning on \(x\).
Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
when \(x\) causes \(y\), where
\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
value of \(y_o\) using the measured value of \(x_o\).

We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
and \(cov(g, \hat{y}_o)\).

@@

\[
\begin{aligned}
cov(g, y_o) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.

@@

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1). In the presence of measurement error in \(x\) and \(y\),
however, Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_o} & > \rho_{g, y_o} \\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
\end{aligned}
\]

@@

The Steiger test is susceptible to inferring the wrong direction of
causality under some measurement error parameters. For an observed
association between the instrument \(g\) and the measured exposure
\(x_o\), \(\rho_{g,x_o}\), the association between \(g\) and the true
exposure (without measurement error) \(x\),
\(\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}\) where the bounds
\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\) are the theoretical limits of
the correlation between true and measured \(x\). The same logic applies
to the association between the instrument and the outcome variable.

For the values \(\rho_{g,x_o} = 0.3\) and \(\rho_{g,y_o} = 0.05\), the
possible true \(\rho_{g,x}\) and \(\rho_{g,y}\) enumerated for all
possible values of measurement error can be visualised as so:

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}

Here, where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error
range in \(x_o\) and \(y_o\) will not change the qualitative inference
of the direction of causality. To evaluate the sensitivity, \(S\) of the
Steiger test to measurement error, the objective is to calculate the
proportion of the total absolute volume of this function that is above
the \(z = 0\) plane,

\[
S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
\]

\(V_{z}\), the total volume of the function, can be obtained
analytically by solving the double integral:

\[
\begin{aligned}
V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
\end{aligned}
\]

\(V_{z>0}\), the proportion of the volume that lies above the \(z=0\)
plane, can also be obtained analytically. The region of this volume is
bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) where
\(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,

\[
\begin{aligned}
V_{z>0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
\]

\(S\) can take values from 0 to 0.5, where low values suggest low
sensitivity of the Steiger test to measurement error. For the values
used to plot the function above, \(\rho_{g,x_o} = 0.3\) and
\(\rho_{g,y_o} = 0.05\), we obtain \(S = 0.032\). Hence, the majority of
the parameter space of measurement error will not change the inferred
causal direction of \(x \rightarrow y\) for these values.
